\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-27 paper/examples.tex
paper: example (phillip feedback)

\section{Motivating Example}
In this section, we present a common scenario where a scientists
has to perform analysis on real-world data.

\subsection{Analysis with missing data}
An epidemiologist at the CDC is tasked with an 
exploratory analysis of data collected from individuals across
a battery of exams. In particular, she is interested in exploring
the relationship between income and the immune system,
controlling for factors such as weight and gender.

%While careful modeling will be necessary before putting
%results into the final research paper,

@@

Before she can perform her queries, the researcher must develop a strategy
for handling the missing values. She currently has two options:
1) she could drop records with missing values in relevant fields,
2) she could use a standard data imputation package on all of her data. Both of these
approaches have significant drawbacks. For the query pictured in ~\Cref{fig:example-query}, 
(1) drops 1492 potentially relevant tuples,
while her previous experiences with (2) have proven be expensive. In a previous project, she
worked with the American Community Survey data (671,153 rows), and performing imputation on the whole
dataset took around 75 minutes\todo{Here, we can reference the number from our own experiment (done), or the number from Akande et al: ``performing imputation on just a 1.5 percent random sample took 45 minutes''}. The analyst needs a more complete picture
of the data, so (1) is insufficient, and she is under time pressure, so (2)
is infeasible.

@@

\begin{figure}
\begin{lstlisting}[language=SQL]
SELECT income, AVG(white_blood_cell_ct)
FROM demo, exams, labs
WHERE gender = 2 AND 
      weight >= 120 AND
      demo.id = exams.id AND 
      exams.id = labs.id
GROUP BY demo.income
\end{lstlisting}
\caption{A typical public health query on CDC's NHANES data}
\label{fig:example-query}
\end{figure}

\begin{figure}
  \Tree
  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
      [.\colorbox{pink}{$\mu_{\text{demo.income}}$}
        [.$\bowtie_{\text{exams.id} = \text{demo.id}}$
          [.\colorbox{pink}{$\mu_{\text{labs.white\_blood\_cell\_ct}}$}
            [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
              [.$\sigma_{\text{exams.weight} \geq 120}$ 
                [.\colorbox{pink}{$\mu_{\text{exams.weight}}$} exams ] 
              ] 
              labs 
            ]
          ]
        [.$\sigma_{\text{demo.gender} = 2}$ demo ]
      ] 
    ] 
  ] 
  ]
\vspace{0.5\baselineskip}
\caption{A quality-optimized plan for the query in~\Cref{fig:example-query}. $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate. $\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}).}
\label{fig:quality-plan}
\end{figure}

\subsection{Planning with \ProjectName{}}
The multiple possible locations for performing imputation leads to a  search space contains plans of varying performance and quality of results. The user can influence the final plan selected by \ProjectName{}
through a parameter $\alpha \in [0, 1]$, where low $\alpha$ 
prioritizes the quality of the query results and high $\alpha$ prioritizes performance.

@@

The quality-optimized plan (\Cref{fig:quality-plan}) uses the impute operation $\mu$, which employs regression model to fill in missing values rather than dropping tuples.
It waits to impute \verb|demo.income| until after the final join has taken place, but other imputations take place earlier on in the plan, some before filtering and join operations.
Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
Meanwhile, the performance-optimized plan (\Cref{fig:fast-plan}) uses the drop operation $\delta$ instead of filling in missing values.
However, it is a significant improvement over dropping all tuples with missing data, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan. In both cases, only performing
imputation on the necessary data yields a query that is much faster
than performing imputation on the whole dataset and then executing
the query.

\begin{figure}
  \Tree
  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
      [.\colorbox{pink}{$\mu_{\text{demo.income}}$}
        [.$\bowtie_{\text{exams.id} = \text{demo.id}}$
          [.\colorbox{pink}{$\mu_{\text{labs.white\_blood\_cell\_ct}}$}
            [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
              [.$\sigma_{\text{exams.weight} \geq 120}$ 
                [.\colorbox{pink}{$\mu_{\text{exams.weight}}$} exams ] 
              ] 
              labs 
            ]
          ]
        [.$\sigma_{\text{demo.gender} = 2}$ demo ]
      ] 
    ] 
  ] 
  ]
\caption{A performance-optimized plan for the query in~\Cref{fig:example-query}. $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate. $\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}).}
\label{fig:fast-plan}
\end{figure}

\subsection{\ProjectName{} Workflow}
Now that the analyst has \ProjectName{}, she can explore
the dataset using SQL. For each query, she repeatedly chooses a value
of $\alpha$ and runs the query until she is unsatisfied with the runtime.
Rather than commit \emph{a priori} to a runtime bound for her query,
which may be unsatisfiable, this iterative approach gives her 
immediate feedback.

>>>>

\section{Motivating Example}
An epidemiologist at the CDC is tasked with an 
exploratory analysis of data collected from individuals across
a battery of exams. In particular, she is interested in exploring
the relationship between income and the immune system,
controlling for factors such as weight and gender.

@@

Before she can perform her queries, the epidemiologist must develop a strategy
for handling missing values. She currently has two options:
1) she could drop records with missing values in relevant fields,
2) she could use a standard data imputation package on all of her data. Both of these
approaches have significant drawbacks. For the query pictured in ~\Cref{fig:example-query}, 
(1) drops 1492 potentially relevant tuples,
while her previous experiences with (2) have proven be expensive. The epidemiologist needs a more complete picture
of the data, so (1) is insufficient, and she is under time pressure, so (2)
is infeasible.

@@

\begin{figure}
\begin{lstlisting}[language=SQL]
SELECT income, AVG(white_blood_cell_ct)
FROM demo, exams, labs
WHERE gender = 2 AND 
      weight >= 120 AND
      demo.id = exams.id AND 
      exams.id = labs.id
GROUP BY demo.income
\end{lstlisting}
\caption{A typical public health query on CDC's NHANES data.}
\label{fig:example-query}
\end{figure}

\begin{figure}
  \Tree
  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
      [.\colorbox{pink}{$\mu_{\text{demo.income}}$}
        [.$\bowtie_{\text{exams.id} = \text{demo.id}}$
          [.\colorbox{pink}{$\mu_{\text{labs.white\_blood\_cell\_ct}}$}
            [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
              [.$\sigma_{\text{exams.weight} \geq 120}$ 
                [.\colorbox{pink}{$\mu_{\text{exams.weight}}$} exams ] 
              ] 
              labs 
            ]
          ]
        [.$\sigma_{\text{demo.gender} = 2}$ demo ]
      ] 
    ] 
  ] 
  ]
\vspace{0.5\baselineskip}
\caption{A quality-optimized plan for the query in~\Cref{fig:example-query}. The operators $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate. $\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}).}
\label{fig:quality-plan}
\end{figure}

\subsection{Planning with \ProjectName{}}
The multiple possible locations for performing imputation leads to a  search space contains plans of varying performance and quality of results. The user can influence the final plan selected by \ProjectName{}
through a tradeoff parameter $\alpha \in [0, 1]$, where low $\alpha$ 
prioritizes the quality of the query results and high $\alpha$ prioritizes performance.

@@

\Cref{fig:quality-plan} shows a quality-optimized plan that uses the impute operation $\mu$, which employs regression model to fill in missing values rather than dropping tuples.
It waits to impute \verb|demo.income| until after the final join has taken place, but other imputations take place earlier on in the plan, some before filtering and join operations.
Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
Meanwhile, \Cref{fig:fast-plan}'s performance-optimized plan uses the drop operation $\delta$ instead of filling in missing values.
However, it is a significant improvement over dropping all tuples with missing data, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan. In both cases, only performing
imputation on the necessary data yields a query that is much faster
than performing imputation on the whole dataset and then executing
the query.

\begin{figure}
  \Tree
  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
      [.\colorbox{pink}{$\mu_{\text{demo.income}}$}
        [.$\bowtie_{\text{exams.id} = \text{demo.id}}$
          [.\colorbox{pink}{$\mu_{\text{labs.white\_blood\_cell\_ct}}$}
            [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
              [.$\sigma_{\text{exams.weight} \geq 120}$ 
                [.\colorbox{pink}{$\mu_{\text{exams.weight}}$} exams ] 
              ] 
              labs 
            ]
          ]
        [.$\sigma_{\text{demo.gender} = 2}$ demo ]
      ] 
    ] 
  ] 
  ]
\caption{A performance-optimized plan for the query in~\Cref{fig:example-query}.The operators $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate. $\mu$ and $\delta$ are specialized imputation operators (\Cref{sec:operators}).}
\label{fig:fast-plan}
\end{figure}

\subsection{\ProjectName{} Workflow}
Now that the epidemiologist has \ProjectName{}, she can explore
the dataset using SQL. For each query, she repeatedly chooses a value
for $\alpha$ and runs the query until she is unsatisfied with the runtime.
Rather than commit \emph{a priori} to a runtime bound for her query,
which may be unsatisfiable, this iterative approach gives her 
immediate feedback.

### 2017-02-27 paper/introduction.tex
paper: example (phillip feedback)

>>>>

\todobox{find appropriate location for this}
{In a previous project, she
worked with the American Community Survey data (671,153 rows), and performing imputation on the whole
dataset took around 75 minutes\todo{Here, we can reference the number from our own experiment (done), or the number from Akande et al: ``performing imputation on just a 1.5 percent random sample took 45 minutes''}.}
